---
title: "Delphi tooling overview"
author: "Daniel J. McDonald"
date: "15 September 2022"
output:
  xaringan::moon_reader:
    css: [src/xaringan-themer.css, src/slides-style.css]
    nature:
      beforeInit: ["src/macros.js"]
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
    seal: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(fontawesome)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  dev = "svg",
  fig.path = "gfx/",
  fig.align = 'center',
  fig.width=9, fig.height=3.5, fig.retina=3,
  fig.showtext = TRUE,
  out.width = "100%",
  cache = TRUE,
  autodep = TRUE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
theme_set(theme_bw(base_size = 14))
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
secondary = "#e98a15"
primary = "#2c365e"
tertiary = "#ffa319"
fourth_color = "#DB0B5B"

style_duo_accent(
  primary_color      = primary,  #"#002145", # UBC primary
  secondary_color    = secondary,  #"6EC4E8", # UBC secondary 4
  header_font_google = google_font("EB Garamond"),
  text_font_google = google_font("Open Sans"),
  code_font_google = google_font("Fira Mono"),
  text_color = primary,
  table_row_even_background_color = lighten_color(primary, 0.8),
  colors = c(
    tertiary = tertiary, fourth_color = fourth_color,
    light_pri = lighten_color(primary, 0.8),
    light_sec = lighten_color(secondary, 0.8),
    light_ter = lighten_color(tertiary, 0.8),
    light_fou = lighten_color(fourth_color, 0.8)
    ),
  outfile = here::here("src/xaringan-themer.css")
)
# theme_set(theme_xaringan())
```

```{r cover, include=FALSE, fig.width=16, fig.height=4}
epipredict::case_death_rate_subset %>%
  filter(case_rate > 1) %>%
  ggplot(aes(time_value, case_rate, colour = geo_value)) +
  geom_line() +
  theme_void() +
  scale_color_viridis_d(option = "C") +
  scale_x_continuous(expand = expansion()) +
  scale_y_log10(expand = expansion(), limits = c(1, NA)) +
  theme(legend.position = "none",
        plot.background = element_rect(fill = primary))

```

layout: true

<div class="my-footer"><span><a href="https://dajmcdon.github.io/delphi-research-tooling" style="color:white">dajmcdon.github.io/delphi-research-tooling</a></span></div> 

---
background-image: url("gfx/cover-1.svg")
background-size: contain
background-position: top

<br/><br/><br/><br/><br/><br/><br/><br/><br/><br/>



.center[#The epiprocess and epipredict R packages]
.center[##Recent tooling work]


.pull-left[
###Daniel J. McDonald
#### 15 September 2022
#### Delphi Research Meeting
]


---

## Delphi's new epiverse

* `{epidatr}` &mdash; R (and python) packages for pulling data from the Delphi API
    - Subsumes the `covidcast` package.
    
    - Access to all the Delphi signals.

* `{epiprocess}` &mdash; R package for processing epi data

* `{epipredict}` &mdash; R package for creating predictive models using the `{epiprocess}` framework


### Credit to: 

Nat, Logan, Sam, Dmitry, Ryan, Ken, Chloe, Maggie, Rachel, (others I'm missing)

---

## `{epiprocess}`

<https://cmu-delphi.github.io/epiprocess/>

1. Data types

    * `epi_df` &mdash; basically a `data.frame` but with important meta information
        1. `as_of` tag to denote the data vintage
        2. time type
        3. geo type
        4. additional keys (e.g., age, gender, race)
    * `epi_archive` &mdash; a collection of `epi_df`s of different vintages
        * But stored so as to eliminate redundancies
        * Allows for important operations (filling, merging, snapshots,  )
        
2. Fundamental functionality &mdash; the `slide()`

    * Rolling correlations
    * Moving averages
    * outlier correction
    * arbitrary functions
    
---

## Example of `slide()`

```{r epidatr-ex, include=FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
library(tidyr)
x <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_7dav_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210601, 20220901),
  geo_values = "ga,pa") %>%
  fetch_tbl() %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

```{r, echo=TRUE}
x <- x %>%  # covid cases in 2 locations downloaded with `{epidatr}`
  group_by(geo_value) %>% 
  mutate(smooth_spline = growth_rate(time_value, cases, method = "smooth_spline"),
         trend_filter = growth_rate(time_value, cases, method = "trend_filter"))
```

```{r, echo=FALSE}
x %>%
  select(geo_value, time_value, smooth_spline, trend_filter) %>%
  pivot_longer(smooth_spline:trend_filter) %>%
  ggplot(aes(x = time_value, y = value)) + 
  geom_line(aes(col = name)) + 
  scale_color_manual(values = c(3,6)) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 2) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "Method") +
  theme(legend.position = "bottom")
```

---

## Examining an `epi_archive`

```{r, echo=FALSE}
x <- archive_cases_dv_subset
x_latest <- epix_as_of(x, max_version = max(x$DT$version))
self_max = max(x$DT$version)
versions = seq(as.Date("2020-06-01"), self_max - 1, by = "1 month")
snapshots <- map_dfr(versions, function(v) { 
  epix_as_of(x, max_version = v) %>% mutate(version = v)}) %>%
  bind_rows(x_latest %>% mutate(version = self_max)) %>%
  mutate(latest = version == self_max) %>%
  filter(geo_value %in% c("ca", "fl"))

ggplot(snapshots %>% filter(!latest),
            aes(x = time_value, y = percent_cli)) +  
  geom_line(aes(color = factor(version))) + 
  geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
  facet_wrap(~ geo_value, scales = "free_y", nrow = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "% of doctor's visits with CLI") + 
  theme(legend.position = "none") +
  geom_line(data = snapshots %>% filter(latest),
               aes(x = time_value, y = percent_cli), 
               inherit.aes = FALSE, color = "black")
library(workflows)
```

* Easy to inspect revision behaviour
* Can use it for pseudo-prospective forecasts
* Compare to our procedure for the PNAS article &mdash; store every version as a `.csv` and load them individually

---
class: middle, center, inverse

# `{epipredict}`

<https://cmu-delphi.github.io/epipredict>

---

## Goals

1.  A set of basic, easy-to-use forecasters that work out of the box.
    You should be able to do a reasonably limited amount of
    customization on them. 
    -   Baseline flat-line forecaster
    -   Autoregressive forecaster
    -   Autoregressive classifier 
    
2.  A .secondary[framework] for creating custom forecasters out of modular
    components. 
    

### Users

.secondary[Basic.] Has data, calls forecaster with default arguments.

.secondary[Intermediate.] Wants to examine changes to the arguments, take advantage of built in flexibility.

.secondary[Advanced.] Wants to write their own forecasters. Maybe willing to build up from some components that we write.

---

## Basic autoregressive forecaster

* Predict `deaths` 1-week ahead on 0, 7, 14 day lags of cases and deaths. 
* Use `lm`, produce intervals with residual quantiles.

```{r, echo = TRUE, warning=FALSE}
library(epipredict)
jhu <- case_death_rate_subset %>% filter(time_value > max(time_value) - 60)
canned <- arx_forecaster(jhu, outcome = "death_rate", 
                         predictors = c("case_rate", "death_rate"))
canned$predictions
```

---

```{r, echo = TRUE}
minor_changes <- arx_forecaster(
  jhu, outcome = "death_rate", predictors = c("case_rate", "death_rate"),
  args_list = arx_args_list(
    ahead = 14, lags = list(c(0:4, 7, 14), c(0, 7, 14)),
    levels = c(0.05, 1:9/10, 0.95),
    quantile_by_key = "geo_value"
  )
)
minor_changes$predictions
```


---

### Use random forests instead

```{r, echo = TRUE}
rf <- arx_forecaster(
  jhu, outcome = "death_rate", predictors = c("case_rate", "death_rate"),
  trainer = parsnip::rand_forest(mode = "regression"), # use ranger
  args_list = arx_args_list(
    ahead = 14, lags = list(c(0:4, 7, 14), c(0, 7, 14)),
    levels = c(0.05, 1:9/10, 0.95),
    quantile_by_key = "geo_value"
  )
)
```


### Or boosting

```{r, echo=TRUE}
xgb <- arx_forecaster(
  jhu, outcome = "death_rate", predictors = c("case_rate", "death_rate"),
  trainer = parsnip::boost_tree(mode = "regression", trees = 20), # use xgboost
  args_list = arx_args_list(
    ahead = 14, lags = list(c(0:4, 7, 14), c(0, 7, 14)),
    levels = c(0.05, 1:9/10, 0.95),
    quantile_by_key = "geo_value"
  )
) 
```

---

## Examine the model

```{r, echo=TRUE}
lmobj <- extract_fit_engine(canned$epi_workflow)
summary(lmobj)
```

---
class: middle, center


Switch to the vignettes:

<https://cmu-delphi.github.io/epipredict>